Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-16 paper-1/manuscript.tex
Converted citations

@@

>>>>

Deviance detection (DD) is the ability of neural systems to identify statistically surprising sensory inputs or violations of established regularities. It is a critical component of any signal processing cascade and is closely related to the concept of prediction errors in the predictive processing framework \citep{Bendixen2012-nx, Khouri2015-gr, Carbajal2018-sd, Fong2020-em}, which has been suggested to underlie most of cognition \citep{Friston2005-jz, Clark2015-gl}. DD has been studied widely across sensory systems including the auditory \citep{Nelken2014-wr, Escera2014-tv, Carbajal2018-sd}, visual \citep{Winkler2012-pr, Pazo-Alvarez2003-kv}, somatosensory \citep{Naatanen2009-jx} domains and beyond. Impairments in DD have been reported in various disorders \citep{Fong2020-em}, including autism \citep{Schwartz2018-bc, Goris2018-bv, Hudac2018-jl, Vlaskamp2017-vs, Chen2020-lu}, schizophrenia \citep{Koshiyama2020-ca, Kim2020-fb, Salisbury2020-sd, Tada2019-lj}, ADHD \citep{Kim2021-uj, Hsieh2021-ti, Lee2020-np} and others, highlighting the need to better understand its mechanistic underpinnings.

Studies of DD must carefully distinguish between repetition suppression and true deviance detection \citep{Ross2020-qf}. The former refers to the reduction of the response to a frequently presented stimulus and can be explained by any adaptive mechanism \citep{May2010-qn, Garagnani2011-eu}. The latter refers to the enhancement of the response to an infrequent stimulus when it is statistically surprising, e.g., due to interrupting a series of regular stimuli, compared to an unsurprising control condition. Here, we study the latter, true DD, which has been suggested to rely on a comparison between prediction signals and received inputs \citep{Parras2017-fp, Carbajal2018-sd, Ross2020-qf}.

Existing computational models of DD place their primary focus on synaptic plasticity as the causal mechanism. While models using long-term plasticity have been proposed \citep{Wacongne2012-ah, Hertag2020-kc}, the rapid establishment of DD in vivo \citep{Taaseh2011-gg} suggests that fast-acting short-term plasticity mechanisms play a key role. Several models using short-term synaptic plasticity have been proposed \citep{Mill2011-ah, May2015-lt, Yarden2017-eh}, finding good agreement with physiological data. However, short-term plasticity also occurs in neuronal excitability, with many neurons exhibiting an experience-dependent transient reduction in intrinsic neuronal sensitivity independent of synaptic function \citep{Sanchez-Vives2000-df, Henze2001-xd, Sanchez-Aguilera2014-fd}. Despite operating at similar time scales to short-term synaptic plasticity, the effects of intrinsic neuronal plasticity on DD is unexplored.  Previous modeling work has only used neuronal plasticity as a side effect of the chosen neuron model \citep{Mill2011-ah} or rejected it a priori as incapable of producing true DD \citep{Yarden2017-eh}. One exception is the work by \cite{Garagnani2011-eu}, which showed that neuronal plasticity in a multi-layer firing rate model could give rise to greater responses to infrequent, surprising stimuli, but did not distinguish between true DD and repetition suppression. Furthermore, it remains unknown how short-term synaptic and neuronal plasticity, which modulate the input to and the output from a neuron, respectively, work together in a network for DD.

Here, we model DD using a network model of spiking neurons with threshold adaptation and short-term synaptic depression, pursuing two aims: Firstly, we aim to show that neuronal adaptation can induce true DD in much the same way as synaptic depression and investigate the mechanistic basis of this effect in detail. Secondly, we aim to investigate the interactions between neuronal and synaptic short-term plasticity in the context of DD. We show that the two mechanisms work similarly and can synergize to enhance their effect. Our results widen the scope of possible mechanisms of DD and highlight the need to consider interactions between various forms of plasticity.

@@

\subsection{Deviance detection emerges from learning regularities}

In this study, we showed that non-synaptic neuronal adaptation can induce true deviance detection in a randomly wired, but spatially organized network of simple leaky integrate-and-fire neurons. Neuronal adaptation achieves this, as illustrated in Figure 10a, by suppressing the response of neurons local to the frequently stimulated site. This drastically reduces the response to frequent stimulations at an early stage of signal propagation through the network, allowing the more distant parts of the network to recover from adaptation and respond vigorously to infrequent deviant stimuli. By contrast, a more varied paradigm including inputs at multiple stimulation sites allows local suppression to partly recover between stimulus presentations, resulting in a greater response through greater recruitment of the wider network. As a result, the network becomes more adapted, diminishing the response to each individual stimulus. Deviance detection, then, is a side effect of learning and adapting to the regularity of the frequently presented stimulus, which leaves the wider network less adapted and more responsive than usual.

\subsection{Biological basis of neuronal threshold adaptation}

Adaptation of the neuronal firing threshold to past activity or membrane depolarization has been demonstrated at a wide variety of timescales ranging from milliseconds \citep{Henze2001-xd, Badel2008-ld, Higgs2011-cy}, supporting precise spike timing \citep{Azouz2000-vh, Huang2016-ch}, to days \citep{OLeary2010-ar}, supporting activity homeostasis. History dependence on a timescale of hundreds of milliseconds to several seconds, as used in our work, has been described in various systems, including early sensory areas \citep{Baccus2002-jb, Wark2009-sf} and later processing stages up to primary sensory cortices and beyond \citep{Spain1991-fg, Sanchez-Vives2000-df, Henze2001-xd, La_Camera2006-dz, Abolafia2011-ig}. Commonly described as spike-frequency adaptation (SFA), this intrinsic feature of neurons supports contrast adaptation \citep{Sanchez-Vives2000-df, Wang2003-vv}, enhances the signal-to-noise ratio \citep{Wark2007-rx, Chacron2007-ak}, and improves signal coding properties of the wider network \citep{Prescott2008-oa, Bohte2012-gk, Lee2023-bu}. SFA can be induced by a number of mutually independent mechanisms, including a buildup of inactivation in voltage-gated sodium channels \citep{Fleidervish1996-ly, Miles2005-nz}, and increased gating of calcium- and sodium-dependent potassium channels \citep{Madison1984-zc, Schwindt1989-lv, Sanchez-Vives2000-rv, Wang2003-vv, Faber2005-vc} and calcium-gated chloride channels \citep{Ha2016-lc, Ha2017-bh}. Perhaps due to this variety in underlying mechanisms, how to model SFA in integrate-and-fire neurons has been debated, and models have been proposed using both hyperpolarizing currents \citep{Izhikevich2003-ty, Brette2005-ci} and adaptive thresholds \citep{Segundo1968-eu, Chacron2003-oz, Kobayashi2009-av}. While there are subtle differences between current- and threshold-based models, and theoretical work has shown current-based approaches to be somewhat superior in capturing firing rate dynamics \citep{Liu2001-yw, Benda2010-tz}, recent large-scale model fitting studies have done very well with adaptive threshold models similar to that used here \citep{Pozzorini2015-ei, Teeter2018-iz}.

\subsection{Threshold adaptation supports deviance detection}

It has been suggested that non-synaptic adaptation such as TA lacks stimulus specificity and is therefore unable to support DD \citep{Duque2015-eu, Yarden2017-eh}. In addition, a frequent argument made is that the various timescales involved in stimulus-specific adaptation (SSA) in an oddball context \citep{Ulanovsky2004-nf} is best mirrored in the similar range of synaptic plasticity timescales \citep{Tsodyks1997-qt, Varela1997-nr}. Yet, in the auditory system, where DD has been studied most intensely, there is intrinsic adaptation at all stages, from the auditory nerve \citep{Yates1983-uu, Westerman1984-ge} through subcortical stations \citep{Ingham2004-wc} to auditory cortex \citep{Abolafia2011-ig}. The timescales involved exhibit a hierarchy, going from tens to hundreds of milliseconds in the sensory and subcortical neurons to several seconds in cortex, a range likewise very similar to that seen in oddball SSA \citep{Ulanovsky2004-nf}. Furthermore, this hierarchy is accompanied by a similar hierarchy in SSA, with thalamus and inferior colliculus being sensitive to shorter stimulus intervals than cortex \citep{Perez-Gonzalez2014-pm}. Finally, it has been shown that combining intrinsic adaptation with STD imbues networks with exquisite sensitivity to changes in the input signal \citep{Puccini2006-tm}. Having shown that TA can support true DD in an oddball context, we anticipate that future models of DD and its analogues will consider including TA or other forms of intrinsic plasticity in addition to synaptic depression.

\subsection{Comparison to other models of DD}

In many ways, our model closely resembles the more elaborate and explicitly cortical model proposed by \cite{Yarden2017-eh}, which uses short-term synaptic depression as its adaptation mechanism. Local suppression is implemented in their model in synaptic depletion in the cortical column sensitive to the target stimulus and is greatest for the standard and least for the deviant, with the many-standards paradigm consistently evoking significant activity across the wider network and inducing an elevated background suppression. In addition, both their model and earlier experimental results \citep{Hershenhoren2014-bb} show occasional large responses to standard, which we have also observed anecdotally. Notably, however, our model operates without explicit compartmentalization into columns. Furthermore, \cite{Yarden2017-eh} report a fine balance of parameter tuning, outside of which their model ceases to operate as intended and does not detect deviant inputs. While we did not explore parameter sensitivity, we take the very minimal tuning and exclusion criteria necessary in our model to indicate a certain level of robustness. With looser connectivity requirements and a weak dependence on parameter settings, our model is applicable to a wider range of neural architectures, including non-cortical circuitry. On the flipside, the nature of the structural constraint imposed makes it difficult to judge whether DD arises as a result of lateral inhibition or adaptation of lateral excitation as in \citep{Yarden2017-eh, Hertag2020-kc, Park2020-oa} or, conversely, as a result of adaptation in feed-forward connections as in \cite{Mill2011-ah}. We speculate that the robustness of our model derives at least in part from this ambiguity: Like in the model by \cite{May2015-lt}, TA reduces all activity propagation, regardless of its destination, thus reducing the impact of structural properties.

In contrast to the predominant thinking in the field (see e.g. \cite{Ross2020-qf} for an extensive review), our work suggests that repetition suppression, which Ross and Hamm call SSA, and true DD can be mediated by a single mechanism, and that DD in particular does not require complex connectivity, as noted above. Indeed, while increased responses to deviant stimuli have been reported in cortex of many mammals, including humans \citep{Tada2019-lj}, other primates \citep{Takaura2016-eh}, and rats \citep{Taaseh2011-gg, Shiramatsu2013-qz, Harms2014-ah}, recent work has shown true DD even in dissociated neuronal culture \citep{Kubota2021-dx}, where connectivity is assumed to be comparatively simple, much like in our model. Rather than invalidating work dissecting the circuit-level mechanisms of cortical DD \citep{Hamm2016-lm, Attinger2017-dw, Yarden2022-yy}, our model provides a possible explanation for the demonstrated presence of DD in subcortical sensory areas \citep{Parras2017-fp} and suggests that what Parras and colleagues identify as prediction errors may in fact persist in the absence of top-down predictions.

\subsection{Adaptation supports subcortical DD}

There is indeed some evidence for this. If feedback signals within cortex and from cortex to midbrain sensory areas carry predictions that cancel out incoming activation, as is commonly assumed in the predictive coding framework \citep{Rao1999-xc, Friston2005-jz, Carbajal2018-sd}, then we should expect these signals to interfere with adaptation-dependent processing, since neurons silenced by top-down predictions are unable to build up a stimulus history with adaptation or depression. Thus, knocking out predictive feedback should facilitate such adaptation-driven DD, while any DD based on predictive inputs would disappear. Indeed, such differential effects have been reported by \cite{Harms2018-bg}. Under the influence of NMDA blockers, which inhibit long-term plasticity in cortex and therefore prevent the formation of accurate predictive models, late components of DD in an oddball paradigm are reduced, while components within the first 10-20 ms after stimulus onset are enhanced. This suggests that early DD, likely driven by subcortical processing \citep{Grimm2012-tb, Parras2017-fp, Font-Alaminos2020-fs}, may be driven by adaptive mechanisms such as that demonstrated in our work, whereas later, cortical DD derives largely from more sophisticated computations involving explicit prediction and error signals \citep{Schlossmacher2022-zv, Grimm2016-fr}. It would be very interesting to explore this interaction in a hierarchical model that combines the adaptation-based approach in early, subcortical areas with a predictive approach \citep{Wacongne2012-ah, Wacongne2016-ht, Hertag2020-kc} in higher, cortical areas.

\subsection{TA and STD synergize to yield supralinear scaling}

Our second main finding is that synaptic short-term depression, which acts in similar fashion to neuronal adaptation both in terms of how it is induced and in how it affects activity flow in the network, enhances the DD effect of neuronal adaptation, and vice versa. We demonstrated that this synergy is a result of increased local suppression of the frequently presented stimulus by STD, which decreases global suppression by both mechanisms in oddball sequences and enhances the deviant response, as illustrated in Figure 10b. Conversely, without strong local suppression in control sequences, global suppression is increased, leading to smaller individual responses.

This synergy can be understood as a simple nonlinear effect of adaptation. While the effect on local suppression of adding STD is direct and mediated only by STD, the resulting reduction of global suppression in the oddball sequence depends entirely on the pre-existing suppression levels mediated by TA and is indeed counteracted to some extent by STD. We speculate that increasing the strength of TA, rather than adding STD, would have a similar effect. Thus, prior to saturation, each additional unit of adaptation from any sufficiently similar source would lead to increasing gains in DD. While it is a priori unsurprising that short-term plasticity should give rise to nonlinearities, the direction of the nonlinearity typically reflects the direction of plasticity: Presynaptic facilitation leads to supralinear activation of the post-synaptic side, and vice versa, depression leads to sublinear activation \citep{Tsodyks1997-qt, Varela1997-nr}, which is typically reflected also in population-level activity \citep{Tsodyks1998-bq}. The reverse has been shown in the context of inhibition-stabilized networks \citep{Wu2022-qx} as a result of the so-called paradoxical effect that increasing excitatory drive decreases activity \citep{Tsodyks1997-hr, Sanzeni2020-su}. However, we have no reason to believe that our model is inhibition-stabilized and are not aware of other work showing supralinear scaling from depression. It would be interesting to confirm our intuition in the model presented here, as well as in a context of more realistic neural and network dynamics.

In addition, while interactions between different forms of short-term plasticity have been investigated in detail at the level of synaptic processes \citep{Tsodyks1997-qt, Varela1997-nr, Buonomano1998-hf, Hennig2008-sc, Anwar2017-qg, Mondal2022-av}, we are aware of only two studies on the interaction between synaptic and neuronal short-term plasticity, investigating oscillations \citep{Mejias2011-rv} and spike train irregularities \citep{Ileri2015-eu} caused by such interactions, respectively. A principled investigation of the interactions between short-term plasticity mechanisms is missing and would seem to be an important addition to our understanding of neural circuit dynamics.

\subsection{Differences between TA and STD}

In the present work, we chose parameters such that TA had a much stronger effect than STD, particularly with the chosen time scales. We believe that the chosen values fall within a physiologically plausible range, and it was not our intention to skew the results in favor of TA. Indeed, we consider it likely that a more even tuning would yield a model capable of evoking comparable levels of DD with either mechanism, working individually and jointly as demonstrated here.

In one aspect, however, neuronal and synaptic plasticity clearly differ: While TA throttles the output of the affected neurons, STD throttles the input to downstream targets. A re-examination of Figure 10b-c, for example, shows that STD affects more neurons as a direct result. The greater reach of the immediate effect of STD perhaps explains why TA has often been overlooked for deviance detection, besides the mistaken claim of stimulus specificity. We also note that this greater reach may result in greater overlap between the neurons experiencing suppression from several stimulus classes. This, in turn, would mean that STD may facilitate the kinds of interactions between different stimuli that give rise to DD, suggesting its applicability to a wider range of neural architectures and of stimuli than TA.

\subsection{Impact of spontaneous activity}

In comparison to biological sensory processing systems, our model is, of course, very simple. One key limitation is in the activity dynamics: Unlike living systems, our model does not produce any spontaneous activity, and evoked responses die out quickly. In contrast, even dissociated cell cultures show some amount of ongoing activity in absence of stimulation. Increased activity would likely substantially affect our results by preferentially following less adapted pathways and therefore reducing the contrast between adapted and recovered parts of the network. Greater spontaneous activity has indeed been linked to reduced DD in rodent inferior colliculus \citep{Duque2012-vx, Duque2015-eu} and auditory cortex \citep{Klein2014-op}, though sensitivity appears to be considerably less than we would expect in our model. In addition, \cite{Abolafia2011-ig} showed that increased background activity also attenuates the level of adaptation to stimuli in individual neurons.

Relatedly, the lack of spontaneous activity also implies that no explicit predictive signals could possibly be generated in our model. Such signals would be necessary for the network to respond to the omission of expected stimuli, as has been demonstrated in cortex of humans \citep{Lehmann2016-ol, Demarchi2019-ch}, marmosets \citep{Jiang2022-ak}, and rats \citep{Auksztulewicz2022-gq}. However, we are unaware of any work showing conclusively the presence or absence of omission responses in subcortical areas. Evidence from EEG presented by \cite{Lehmann2016-ol} suggests that no such responses may occur, but a finer-grained analysis in animal models would help to further narrow down to what extent the early, subcortical DD processes are driven by prediction and adaptation \citep{Schlossmacher2022-zv}. Along with other considerations outlined above, the absence of evidence of omission responses seems to suggest that our adaptation-based model is a better description of subcortical processing.

\subsection{Early and late response in culture}

In dissociated culture \citep{Kubota2021-dx}, the stimulus response is biphasic, with a peak of directly evoked activity in the first 10 ms, and a longer response lasting up to 200 ms. In the results reported by Kubota and colleagues (Figure 2), DD appears to become evident only in the later response. Since we do not simulate conduction delays, the response in our model is compressed over a much shorter time period. However, we also see a larger response difference in the later stage of the response, as activity passes from the stimulated neurons and their immediate neighborhood into the wider network. It remains to be confirmed whether the distinction between early and late responses in culture is tied to separate neuron populations, as in our model, indicating some amount of linearity of activity propagation, or whether the early and late responses both develop in a distributed manner throughout the network.

\subsection{Conclusion}

As alluded to above, we anticipate that future models of DD and its attendant signatures (SSA, mismatch negativity, etc.) will rely on hierarchical structure, including not just different cortical areas \citep{May2015-lt, Park2020-oa}, but also subcortical areas with their own processing dynamics. Given the evidence presented here, we strongly believe that neuronal adaptation will play a key role in such future models, working alongside synaptic short-term plasticity \citep{Yarden2017-eh} and long-term plasticity \citep{Hertag2020-kc} to better explain the full richness of temporal sensory processing.
